Supplementary material for referee\\MNRAS: MN-15-0213-MJ

### 2015-03-25 response/supplement_response.tex
working on response

%%%% Various options for document class.
%\documentclass[usenatbib, a4paper]{preprint}
\documentclass[preprint, a4paper, 11pt]{aastex}
%%\documentclass[twocolum]{revtex4}
%%\documentclass{report}
%%\documentclass[useAMS,usenatbib, onecolumn]{mn2ejm}

%\usepackage{psfig,morefloats,url}
%use preprint2 for 2 columns paper.

%% declare any packages used
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{color} 
\usepackage{pdfpages}
\usepackage{appendix}
\usepackage{subfigure}
\usepackage{amsmath}
\usepackage{amssymb}

\citestyle{aa}
\addtolength{\topmargin}{-.5in}
%% This command added as margins are wrong in mn2e, it appears. 
%% Not needed for other classes

%% This is the end of the preamble.  Indicate the beginning of the
%% paper itself with \begin{document}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
%% define bibstyle and other definitions
\bibliographystyle{aabib}
%% renew commands
\renewcommand{\labelitemi}{$\bullet$}
\def\py{\textsc{Python}}
\def\tar{\textsc{Tardis} }
\def\cld{\textsc{Cloudy} }
\def\civ{C~\textsc{iv} }
\def\araa{ARAA}
\def\nat{Nature}
\def\apjl{ApJ Letters}
\def\aapr{AAPR}
\def\actaa{ACTAA} 
\def\ssr{SSR}
\def\apj{ApJ}
\def\apss{AP\&SS}
\def\pasp{PASP}
\def\aap{A\&A}
\def\mnras{MNRAS}
\def\aj{AJ}
\def\rmxaa{RMXAA}

%% define some control sequences for lines
\def\heiiopt{He~\textsc{ii}~$\lambda4686$}
\def\heiiuv{He~\textsc{ii}~$\lambda1640$}
\def\heiioptnew{He~\textsc{ii}~$\lambda3202$}
\def\la{Ly$\alpha$}
\def\ha{H$\alpha$}
\def\hb{H$\beta$}
\def\civfull{C~\textsc{iv}~$\lambda1550$}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          TITLE AND AUTHORS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{The Impact of Accretion Disk Winds on the Optical Spectra of Cataclysmic Variables}

\author[Matthews et al.]{J.~H.~Matthews,$^1$ C.~Knigge,$^1$ K.~S.~Long,$^2$ S.~A.~Sim,$^3$ and N.~Higginbottom$^4$
\medskip  
\\$^1$School of Physics and Astronomy, University of Southampton, Highfield, Southampton, SO17 1BJ, UK
\\$^2$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD 21218, USA
\\$^3$School of Mathematics and Physics, Queens University Belfast, University Road, Belfast, BT7 1NN, Northern Ireland, UK
\\$^4$Department of Astronomy, University of Las Vegas, Las Vegas, NV 89119, USA}

%%\author{J. H. Matthews, C. Knigge, K. S. Long, S. A. Sim \& N. Higginbottom}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          ABSTRACT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
% Many high-state non-magnetic cataclysmic variables (CVs) exhibit
% blue-shifted absorption or P-Cygni profiles associated with ultraviolet
% (UV) resonance lines, such as C~{\sc iv}~$\lambda1550$. These features
% imply the existence of powerful accretion disk winds in
% CVs. Here, we use an improved version of our Monte Carlo ionization
% and radiative transfer code to investigate whether disk wind models that
% produce realistic UV line profiles are also likely to generate
% observationally significant recombination line and continuum emission
% in the {\em optical} waveband. 
% We also test whether outflows may be responsible for the single-peaked 
% emission line profiles often seen
% in high-state CVs and for the weakness of the Balmer absorption edge
% (relative to simple models of optically thick accretion disks).
% We find that a standard disk wind model that is successful in
% reproducing the UV spectra of CVs also leaves a noticeable imprint on
% the optical spectrum, particularly for systems viewed at high
% inclination. The strongest optical wind-formed recombination lines are \ha\ and
% \heiiopt, although the outflow also produces emission in the Ly$\alpha$
% and \heiiuv\ UV lines. However, all of the optical lines produced by
% this default wind model are double-peaked, and the effect of the wind
% on the Balmer jump is modest. We demonstrate that a higher-density outflow model 
% displays reasonable verisimilitude with the optical spectrum of RW Trianguli,
% producing all the observed lines. This model
% produces a recombination continuum that is strong enough to fill in the Balmer jump,
% but it also predicts stronger-than-observed collisionally excited UV emission lines. 
% The wind in our revised model also alters the spectral shape of the continuum, 
% and displays a strong inclination dependence in its emission properties. 
% No single-peaked emission is seen, although we observe a 
% narrowing of the double-peaked emission lines from the base of the wind.
% We show that even denser models can produce a single-peaked \ha\ line.
% In this model, the single-peaked line profile is due to the line emission originating from 
% $\sim10^{11}$~cm above the disk plane where rotational velocities are small, 
% and is not due to the radial velocity shear mechanism that
% has previously been proposed. On the basis of our results, we suggest that winds can 
% modify, and perhaps even dominate, the line and continuum emission
% from CVs.
Many high-state non-magnetic cataclysmic variables (CVs) exhibit
blue-shifted absorption or P-Cygni profiles associated with ultraviolet
(UV) resonance lines. These features imply the existence of 
powerful accretion disk winds in CVs. Here, we use our Monte Carlo ionization
and radiative transfer code to investigate whether disk wind models that
produce realistic UV line profiles are also likely to generate
observationally significant recombination line and continuum emission
in the {\em optical} waveband. We also test whether outflows may be responsible for 
the single-peaked emission line profiles often seen in high-state CVs and for the weakness
of the Balmer absorption edge (relative to simple models of optically thick accretion disks).
We find that a standard disk wind model that is successful in
reproducing the UV spectra of CVs also leaves a noticeable imprint on
the optical spectrum, particularly for systems viewed at high
inclination. The strongest optical wind-formed recombination lines are \ha\ and
\heiiopt. We demonstrate that a higher-density outflow model produces 
all the expected H and He lines and produces a recombination continuum that can 
fill in the Balmer jump at high inclinations.
This model displays reasonable verisimilitude with the optical spectrum of RW Trianguli. No single-peaked emission is seen, although we observe a 
narrowing of the double-peaked emission lines from the base of the wind.
Finally, we show that even denser models can produce a single-peaked \ha\ line.
On the basis of our results, we suggest that winds can 
modify, and perhaps even dominate, the line and continuum emission
from CVs.
\end{abstract}

\begin{keywords}
{\sl (stars:)} novae, cataclysmic variables - accretion, accretion discs - stars: winds, outflows - line: profiles - radiative transfer - methods: numerical 
\end{keywords}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          INTRODUCTION
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction} 
\label{sec:intro}

\begin{figure}	%fullpage
\centering
\includegraphics[width=0.7\textwidth]{figures/fig1.eps}
\caption{
Optical spectra of three nova-like variables: 
RW Sex (top; Beuermann et al. 1992),
IX Vel (top middle; A. F. Pala \& B. T. Gaensicke, private communication) 
and RW Tri in and out of eclipse (bottom two panels; Groot et al. 2004).
The data for RW Sex and RW Tri were digitized from the respective publications,
and the IX Vel spectrum was obtained using the XSHOOTER spectrograph 
on the Very Large Telescope on 2014 October 10.
These systems have approximate inclinations of $30^\circ$, $60^\circ$ and $75^\circ$ 
(see section 5.4) respectively. 
The trend of increasing Balmer line emission with inclination can be seen.
In RW Tri strong single-peaked emission in the Balmer lines is seen even
in eclipse, indicating that the lines may be formed in a spatially
extensive disk wind, and there is even a suggestion 
of a (potentially wind-formed) recombination continuum in the eclipsed
spectrum. We have attempted to show each spectrum over a similar dynamic range.
}
\label{novalikes}
\end{figure}

Cataclysmic variables (CVs) are systems in which a white dwarf
accretes matter from a donor star via Roche-lobe overflow. In
non-magnetic systems this accretion is mediated by a Keplerian disk
around the white dwarf (WD). Nova-like variables (NLs) are a subclass
of CVs in which the  disk is always in a relatively
high-accretion-rate state ($\dot{M} \sim
10^{-8}$~M$_{\odot}$~yr$^{-1}$).  This makes NLs an excellent
laboratory for studying the properties of steady-state accretion
disks.

It has been known for a long time that winds emanating from the
accretion disk are important in shaping the ultraviolet (UV) spectra
of high-state CVs \citep{heap1978, greensteinoke1982}. The most spectacular evidence for such
outflows are the P-Cygni-like profiles seen in UV resonance lines such as
\civfull\ (see e.g. Cordova \& Mason
1982\nocite{cordova1982}). Considerable effort has been spent over the
years on understanding and modelling these UV features (e.g. Drew \&
Verbunt 1985\nocite{drewverbunt1985}; Mauche \& Raymond
1987\nocite{maucheraymond1987}; Drew 1987; Shlosman \& Vitello 1993; [hereafter
SV93]\nocite{SV93}; Knigge, Woods \& Drew 1995\nocite{KWD95}; 
Knigge \& Drew 1997\nocite{kd1997}; 
Knigge et al. 1997\nocite{knigge1997}; Long \& Knigge 2002 [hereafter LK02]\nocite{LK02}, 
Noebauer et al. 2010\nocite{noebauer};
Puebla et al. 2011\nocite{puebla2011}). The basic picture emerging from these efforts is
of a slowly accelerating, moderately collimated bipolar
outflow that carries away $\simeq 1\% - 10\%$ of the accreting
material. State-of-the-art simulations of line formation in this type
of disk wind can produce UV line profiles that are remarkably similar
to observations.

Much less is known about the effect of these outflows on the optical
spectra of high-state CVs. These spectra are typically characterized
by H and He emission lines superposed on a blue continuum. In many
cases, and particularly in the SW~Sex subclass of NLs
\citep{HSK86,DR95}, these lines are single-peaked. This is contrary to
theoretical expectations for lines formed in accretion disks, which
are predicted to be double-peaked \citep{smak1981, hornemarsh1986}. 
{\em Low-state} CVs (dwarf novae in quiescence) do, in fact,
exhibit such double-peaked lines \citep{marshhorne1990}.

Murray \& Chiang (1996, 1997; hereafter referred to collectively as MC96)\nocite{MC96, MC97} 
have shown that the presence of disk winds may
offer a natural explanation for the single-peaked optical emission lines in
high-state CVs, since they can strongly affect the radiative transfer
of line photons. Strong support for a significant wind contribution to the
optical emission lines comes from observations of eclipsing
systems. There, the single-peaked lines are often only weakly
eclipsed, and a significant fraction of the line flux remains visible
even near mid-eclipse \citep[e.g.][]{baptista2000,groot2004}. 
This points to line formation in a spatially
extended region, such as a disk wind (see Fig.~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence for a wind contribution to the optical lines comes
from isolated observations of P-Cygni-like line profiles even in optical
lines, such as \ha\ and He \textsc{i} $\lambda5876$ \citep{patterson1996, RN98, kafka2004}.

Could disk winds also have an impact on the UV/optical {\em continuum}
of high-state CVs? This continuum is usually thought to be dominated
by the accretion disk and modelled by splitting the disk into
a set of concentric, optically thick, non-interacting annuli following
the standard $T_{eff}(R) \propto R^{-3/4}$ radial temperature
distribution \citep{shakurasunyaev1973}. In such
models, each annulus is taken to emit either as a blackbody or,
perhaps more realistically, as a stellar/disk atmosphere model
\citep{Schwarzenberg-Czerny1977,wade1984,wade1988}.
In the latter case, the local surface gravity, $\log{g}(R)$, is
assumed to be set solely by the accreting WD, since self-gravity is
negligible in CV disks.

Attempts to fit the observed spectral energy distributions (SEDs) of
high-state CVs with such models have met with mixed success. In
particular, the SEDs predicted by most stellar/disk atmosphere models 
are too blue in the UV \citep{wade1988,long1991,long1994,knigge1998} and exhibit
stronger-than-observed Balmer jumps in absorption 
\citep{wade1984,haug1987,ladous1989b,knigge1998}. One possible
explanation for these problems is that these models fail to capture
all of the relevant physics. Indeed, it has been argued that a
self-consistent treatment can 
produce better agreement with observational data (e.g. Shaviv et
al. 1991;  but see also Idan et al. 2010).
\nocite{idanshaviv2010} \nocite{shaviv1991}
However, an alternative explanation, suggested by Knigge et al.
(1998b; see also Hassall et al. 1985)\nocite{KLWB98,hassall}, 
is that recombination continuum emission from the base of the 
disk wind might fill in the disk's
Balmer absorption edge and flatten the UV spectrum.

\nocite{groot2004}
\nocite{beuermann1990}
\nocite{beuermann1992}
\nocite{higginbottom2013}

Here, we carry out Monte Carlo radiative transfer simulations in
order to assess the likely impact of accretion disk winds on the
optical spectra of high-state CVs. More specifically, our goal is to
test whether disk winds of the type developed to account for the UV
resonance lines would also naturally produce significant amounts of  
optical line and/or continuum emission. In order to achieve this, we
have implemented the `macro-atom' approach developed by Lucy
(2002, 2003) into the Monte Carlo ionization and radiative transfer
code described by LK02 (a process initiated by Sim et al. 2005; hereafter SDL05). 
With this upgrade, the code is able to deal correctly with processes involving
excited levels, such as the recombination emission produced by CV
winds.

The remainder of this paper is organized as follows. In Section~2, we
briefly describe the code and and the newly implemented macro-atom
approach. In Section~3, we describe the kinematics and geometry of our
disk wind model. 
In Section~4, we present spectra simulated from the benchmark model
employed by LK02, and, in Section~5, we present a revised model
optimized for the optical waveband. In Section~6, we summarize our
findings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          THE CODE, RADIATIVE TRANSFER
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{\sc{python}: A Monte Carlo Ionization and Radiative Transfer Code}

\py\ is a Monte Carlo  ionization and radiative transfer code which
uses the Sobolev approximation to treat line transfer 
\citep[e.g.][]{sobolev1957,sobolev1960,rybickihummer1978}. 
The code has already been described extensively by LK02, SDL05 and Higginbottom et al. (2013; hereafter H13), so here we provide only a brief summary of its operation, 
focusing particularly on new
aspects of our implementation of macro-atoms into the code.

\subsection{Basics}

\py\ operates in two distinct stages. First, the ionization state,
level populations and temperature structure are calculated. This is
done iteratively, by 
propagating several populations of Monte Carlo energy quanta (`photons')
through a model wind. The geometric and kinemetic properties of the
outflow are specified on a pre-defined spatial grid. In each of these
iterations (`ionization cycles'), the code records estimators that 
characterize the radiation field in each grid cell. At the end 
of each ionization cycle, a new electron temperature is calculated
that more closely balances heating and cooling in the 
plasma. The radiative estimators and updated electron
temperature are then used to revise the ionization state of the wind,
and a new ionization cycle is started. The process is repeated until
heating and cooling are balanced throughout the wind.

This converged model is then used as the basis for the second set of
iterations (`spectral cycles'). In these, the emergent spectrum over
the desired spectral range is synthesized by tracking populations of
energy packets through the wind and computing the emergent spectra at
a number of user-specified viewing angles.

\py\ is designed to operate in a number of different
regimes, both in terms of the scale of the system and in terms of the
characteristics of the underlying radiation field.
It was originally developed by LK02 in order to model the UV spectra
of CVs with a simple biconical disk wind model. SDL05
\nocite{simmacro2005} used the code to model Brackett
and Pfund line profiles of H in young-stellar objects (YSOs). As part
of this effort, they implemented a `macro-atom' mode (see below) in
order to correctly treat H recombination lines with
\py. Finally, H13 used \py\ to model broad absorption line (BAL) QSOs. For
this application, an improved treatment of ionization was implemented,
so that the code is now capable of dealing with arbitrary
photo-ionizing SEDs, including non-thermal and multi-component ones.

\subsection{Ionization and Excitation: `Simple Atoms'}
\label{simpleatoms}

Prior to SDL05, the relative ionization fractions for all atomic
species were estimated via the modified Saha equation (Mazzali \&
Lucy 1993)  
\begin{equation}
\frac{n_{j+1} n_e}{n_j} = W [\xi + W(1-\xi)]
\left(\frac{T_e}{T_R}\right)^{1/2}
\left(\frac{n_{j+1}n_e}{n_j}\right)^*_{T_R}. \label{ionization}
\end{equation}
Here, the `starred' term on the right represents abundances computed with
the Saha equation at temperature $T_R$, but using partition functions
from the dilute blackbody approximation. 
$W$ is an effective dilution factor, $\xi$ is the
fraction of recombinations going directly to the ground state, and
$T_R$ and $T_e$ are the radiation and electron temperatures,
respectively. This simple ionization scheme produces reasonable
results when the photoionizing SED can be approximated by a dilute
blackbody. This is the case for high-state CVs. (As noted above, an
improved, but more complex treatment of ionization that is appropriate
for more complex SEDs is described in H13.)

Similarly, the relative excitation fractions within each ionization
stage of a given species were estimated via a modified Boltzmann
equation,
\begin{equation}
\frac{n_{jk}}{n_j} = \frac{W g_k}{z_j(T_R)} \exp(-E_k/kT_R),
\end{equation}
where $n_{jk}$ is the population of level $k$ in ionic stage $j$,
$E_k$ is the energy difference between level $k$ and the ground state,
$g_k$ is the statistical weight of level $k$
and $z_j(T_R)$ is the partition function of ionic stage $j$.

Finally, \py\ modelled all bound-bound processes as transitions
within a simple two-level atom \cite[e.g.][]{mihalas}. This
approach works reasonably well for resonance  
lines, such as \civfull, in which the lower level is the ground state.  
However, it is a poor approximation for many other
transitions, particularly those where the upper level
is primarily populated from above. Thus an improved method for
estimating excited level populations and simulating line transfer is
needed in order to model recombination lines and continua.

\subsection{Ionization and Excitation: Macro-Atoms}

Lucy (2002, 2003\nocite{lucy2002, lucy2003}; hereafter L02, L03) 
has shown that it is possible to calculate the emissivity of a gas in
statistical equilibrium accurately by quantising matter into
`macro-atoms', and radiant and kinetic energy into indivisible energy
packets (r- and k- packets, respectively). His macro-atom scheme
allows for all possible transition paths from a given level and
provides a full non-local thermodynamic equilibrium (NLTE) solution
for the level populations based on Monte Carlo estimators. The macro-atom
technique has already been used to model Wolf-Rayet star
winds \citep{sim2004}, AGN disk winds \citep{simlong2008, tatum2012},
supernovae \citep{kromersim2009, kerzendorfsim} and YSOs (SDL05). A full 
description of the approach can be found in L02 and L03.

Briefly, macro-atom NLTE level populations and ionization fractions
are calculated by solving the statistical equilbrium equations between
each pair of levels. In the framework of the Sobolev escape probability formalism (Rybicki \& Hummer 1978; L02; Sim 2004), 
the bound-bound excitation rate, ${\cal R}_{lu}$, in an ion is given by 
\begin{equation}
{\cal R}_{lu} = B_{lu} n_l J_{est} + C_{lu} n_l n_e,
\end{equation}
where $u$ and $l$ denote the upper and lower levels, $C$ represents the
collisional rate coefficients, and $B$ is the usual Einstein
coefficient. $J_{est}$ is the Monte Carlo estimator for the mean intensity 
impinging on the Sobolev region, weighted by an angle-dependent escape probability, 
given by \citep{sim2004}
\begin{equation}
J_{est} = \frac{c}{4 \pi \nu_0 V} \sum_{photons} w \frac{1 - e^{-\tau_s}}{\tau_s} \frac{1}{dv/ds}.
\end{equation}
Here $w$ is the photon weight (in luminosity units), $\nu_0$
is the line frequency, $dv/ds$ is the velocity gradient and
$\tau_s$ is the Sobolev optical depth.
The sum is over all photons that come into resonance with the line.
The corresponding de-excitation rate is then 
\begin{equation}
{\cal R}_{ul} = \beta_{lu} A_{ul} n_u + B_{ul} n_u J_{est} +
C_{ul} n_u n_e,
\label{eq:nlte_rul}
\end{equation}
where $A$ is the usual Einstein coefficient. 
The quantity $\beta_{lu}$ is the {\em angle-averaged} probability 
that a given line photon will escape the Sobolev region.

In our implementation of the macro-atom approach, we also explicitly
take into account the photoionization and collisional ionization rates
between a lower level, $l$, and the continuum (or, in the case of ions
with more than one bound electron, the ground state of the upper ion),
$\kappa$,
\begin{equation}
{\cal R}_{l \kappa}= n_l \int_{\nu_0}^{\infty} \frac{ 4 \pi J_{\nu}
  \sigma_{\nu}}{h \nu} d\nu + C_{l \kappa} n_l n_e.
\end{equation}
Here, $\sigma_{\nu}$ is the photoionization cross section, and $J_{\nu}$
is the mean intensity. The corresponding recombination rate is given
by 
\begin{equation}
{\cal R}_{\kappa l} = \alpha_{\kappa l} n_{\kappa} n_e + C_{\kappa l}
n_\kappa n_e, \\
\end{equation}
where $\alpha_{\kappa l}$ is the radiative recombination coefficient
to level $l$. This treatment means that radiative and collisional
rates to and from all levels are considered when calculating both the
ionization state and the level populations, although we neglect 
ionization directly to excited levels of the upper ion. The
\cite{vanregemorter} approximation is used for collisional
transitions, and thus we assume that collisions between radiatively
forbidden transitions are not important in determining level
populations. This is a reasonable approximation in the regimes
considered here.

\subsection{Ionization and Excitation: A Hybrid Approach}

SDL05 implemented a macro-atom treatment of H in \py\ and used
this to predict the observable properties of a pure H wind
model for YSOs. Our goal here is to simultaneously model the optical
and ultraviolet spectra of high-state CVs. Since the optical spectra
are dominated by H and He recombination lines, both of these species
need to be treated as macro-atoms. The UV spectra, on the other hand,
are dominated by resonance lines associated with metals. This means we
need to include these species in our models, but they can be treated 
with our (much faster) simple-atom approach. We have therefore
implemented a hybrid ionization and excitation scheme into \py. Any
atomic species can now be treated either in our simple-atom
approximation or with the full macro-atom machinery. In our CV
models, we treat H and He as macro-atoms and all metals as
simple-atoms.

\subsection{Atomic Data}

We generally use the same atomic data as H13, which is an updated
version of that described by LK02. In addition, we follow SDL05 in
treating H as a 20-level atom, where each level is defined by
the principal quantum number, $n$. For the macro-atom treatment of
He, we have added the additional level and line information required 
from \textsc{Topbase} \citep{topbase2005}.  He~\textsc{ii} is treated
in much the same way as H, but with 10 levels. He~\textsc{i} has
larger energy differentials between different l-subshells and triplet
and singlet states. Thus, we still include levels up to $n=10$, but
explicitly treat the $l$ and $s$ sub-orbitals as distinct levels
instead of assuming they are perfectly `l-mixed'. This allows us
to model the singlet and  triplet He~\textsc{i} lines that are ubiquitous
in the optical spectra of CVs \citep[e.g.][]{dhillon1996}.

\subsection{Code Validation and Testing}

\py\ has been tested against a number of radiative transfer and
photoionization codes. LK02 and H13 conducted comparisons of 
ionization balance  with \cld \citep{cloudy2013}, demonstrating
excellent agreement. We have also carried out comparisons
of ionization and spectral synthesis with the supernova code
\textsc{Tardis.} \tar is described by 
\cite{kerzendorfsim}, and the spectral comparisons can be found
therein. For the effort reported here, we have additionally carried
out tests of the macro-atom scheme in \py. Fig.~\ref{tests} shows
two of these tests. In the top panel, we compare the Balmer series 
emissivities as predicted by \py\ in the l-mixed Case~B limit against the
analytical calculations by \cite{seaton1959}. In the bottom panel, we
compare \py\ and \tar predictions of  He \textsc{i} level populations
for a particular test case. Agreement is excellent for both H and He.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/fig_caseb_tardis.eps}
\caption{
{\sl Top Panel:} `Case B' Balmer decrements computed 
with \textsc{Python} compared to analytic calculations
by Seaton (1959). Both calculations are calculated at $T_e=10,000$K.
(see Osterbrock 1989 for a discussion of this commonly used approximation).
{\sl Bottom Panel:}  a comparison of He I level populations (the most complex ion we currently 
treat as a macro-atom) between \py\ and \tar models. 
The calculation is conducted with physical parameters of $n_e=5.96\times10^4$~cm$^{-3}$,
$T_e=30,600$K, $T_R=43,482$K and $W=9.65\times10^{-5}$. 
Considering the two codes use different atomic data and \textsc{Tardis,} unlike \textsc{Python,} currently has a complete treatment of collisions between 
radiatively forbidden transitions, the factor of 
$<2$ agreement is encouraging. 
}
\label{tests}
\end{figure}

\nocite{osterbrock}
\nocite{seaton1959}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          THE MODEL
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Describing the System and its Outflow}

\py\ includes several different kinematic models of accretion disk
winds, as well as different options for describing the physical and
radiative properties of the wind-driving system under
consideration. Most of these features have already been discussed by
LK02 and H13, so below we only briefly recount the key aspects of the
particular system and wind model used in the present study.

\subsection{Wind Geometry and Kinematics}
\label{kinematics}

We adopt the kinematic disk wind model developed by SV93. 
A schematic of this model is shown in
Fig.~\ref{cartoon}. In this parametrization, a smooth, biconical
disk wind emanates from the accretion disk between radii $r_{min}$ and 
$r_{max}$. The covering fraction of the outflow is also controlled by the
inner and outer opening angles of the wind, $\theta_{min}$ and
$\theta_{max}$, and the launch angle of the other streamlines is given
by 
\begin{equation}
\theta(r_0) = \theta_{min} + (\theta_{max} - \theta_{min}) \left(\frac{r_0 - r_{min}}{r_{max} - r_{min}} \right)^{\gamma},
\label{theta}
\end{equation}
where $r_0$ is the launch radius of the streamline.

The poloidal (non-rotational) velocity field of the wind, $v_l$, is given by
\begin{equation}
v_l=v_0+\left[v_{\infty}(r_0)-v_0\right]\frac{\left(l/R_v\right)^{\alpha}}{\left(l/R_v\right)^{\alpha}+1},
\label{v_law}
\end{equation}
where $l$ is the poloidal distance along a particular wind
streamline. The terminal velocity along a streamline, $v_{\infty}$, is
set to a fixed multiple of $v_{esc}$, the escape velocity at the launch
point. The launch velocity from the disk surface, $v_0$, is assumed to
be constant (set to $6$~km~s$^{-1}$). Once the wind is launched, it
accelerates, reaching half of its terminal velocity at $l = R_v$. The
velocity law exponent $\alpha$ controls how quickly the wind
accelerates. Larger values of $\alpha$ cause the main region of 
acceleration to occur close to $R_v$, whereas smaller values
correspond to fast acceleration close to the disk (see
Fig.~\ref{acc_law}). The rotational velocity $v_\phi$ is 
Keplerian at the base of the streamline 
and we assume conservation of specific angular momentum,
such that
\begin{equation}
v_\phi r = v_{k} r_0,
\label{v_law}
\end{equation}
where $v_{k}=(GM_{WD}/r_0)^{1/2}$.

\begin{figure} 
\centering
\includegraphics[width=0.45\textwidth]{figures/fig2_cartoon.eps}
\caption{Cartoon illustrating the geometry and kinematics of the benchmark CV wind model.}
\label{cartoon}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/acc_law.eps}
\caption{
The adopted poloidal velocity law for various values of the
acceleration exponent, $\alpha$.
%%{\bf JM: KSL has suggested we ditch this}
} 
\label{acc_law}
\end{figure}

The density at position $(r,z)$ in the wind, $\rho(r,z)$, is
calculated from the mass continuity equation, yielding
\begin{equation}
\rho(r,z) = \frac{r_0}{r} \frac{dr_0}{dr} \frac{\phi(r_0)}{v_z(r,z)}.
\label{density}
\end{equation}
Here, 
$v_z$ is the vertical velocity component and, following SV93,
$\phi(r_0)$ is the local mass-loss rate per unit area at $r_0$, 
defined as
\begin{equation}
\phi(r_0) \propto \dot{M}_{wind} r_0^\lambda \cos [\theta(r_0)].
\label{density}
\end{equation}
We adopt $\lambda = 0$ and normalize $\phi(r_0)$ by 
matching its integral over both sides of the disk
to the user-specified total mass-loss rate, $\dot{M}_{wind}$.

\subsection{Sources and Sinks of Radiation}
\label{radsources}

The net photon sources in our CV model are the accretion disk, the
WD and, in principle, a boundary layer with user-defined temperature
and luminosity. All of these radiating bodies are taken to be
optically thick, and photons striking them are assumed to be destroyed
instantaneously. The secondary star is not included as a radiation
source, but is included as an occulting body. This allows us to model
eclipses. Finally, emission from the wind itself is also accounted for, but
note that we assume the outflow to be in radiative equilibrium. Thus all
of the heating of the wind, as well as its emission, is ultimately
powered by the radiation field of the net photon sources in the
simulation. In the following sections, we will describe our treatment
of these system components in slightly more detail.

\subsubsection{Accretion Disk}

\py\ has some flexibility when treating the accretion 
disk as a source of photons. The disk is broken down into annuli 
such that each annulus contributes an equal amount to the bolometric
luminosity. We take the disk to be geometrically thin, but optically
thick, and thus adopt the temperature profile of a standard
\cite{shakurasunyaev1973} $\alpha$-disk. An annulus can then
be treated either as a blackbody with the corresponding effective
temperature or as a stellar atmosphere model with the appropriate
surface gravity and effective temperature. Here, we use blackbodies 
during the ionization cycles and to compute our Monte Carlo
estimators. However, during the spectral synthesis stage of the 
simulation we use stellar atmosphere models. This produces more
realistic model spectra and allows us to test if recombination
emission from the wind base can fill in the Balmer jump, which is
always in absorption in these models. Our synthetic stellar atmosphere
spectra are calculated with
\textsc{Synspec}\footnote{http://nova.astro.umd.edu/Synspec43/synspec.html}
from either Kurucz \citep{kurucz1991} atmospheres (for $T_{eff} \leq
50,000$~K) or from \textsc{TLUSTY} \citep{tlusty} models (for $T_{eff} > 50,000$~K).

\subsubsection{White Dwarf}

The WD at the center of the disk is always present as a spherical occulting
body with radius $R_{WD}$ in \py\ CV models, but it can also be included
as a source of radiation. In the models presented here, we treat the
WD as a blackbody radiator with temperature $T_{WD}$ and luminosity
$L_{WD} = 4\pi R_{WD}^2 \sigma T_{WD}^4$.

\subsubsection{Boundary Layer}

It is possible to include radiation from a boundary layer (BL) between
the disk and the WD. In \py, the BL is described as
a blackbody with a user-specified effective temperature and
luminosity. In the models presented here, we have followed LK02 in setting
the BL luminosity to zero. However, we have confirmed that the addition of an isotropic
BL with $L_{BL} = 0.5 L_{acc}$ and temperatures in the range $80~{\rm
kK} \leq T_{BL} \leq 200~{\rm kK}$ would not change any of our main
conclusions.

\subsubsection{Secondary Star}

The donor star is included in the system as a pure radiation sink, 
i.e. it does not emit photons, but absorbs any photons that strike its
surface. The secondary is assumed to be Roche-lobe filling, so its
shape and relative size are defined by setting the mass ratio of the system, 
$q = M_2/M_{WD}$. The inclusion of the donor star as an occulting body
allows us to model eclipses of the disk and the wind. For this
purpose, we assume a circular orbit with a semi-major axis $a$ and 
specify orbital phase such that $\Phi_{orb} = 0$ is the
inferior conjunction of the secondary (i.e. mid-eclipse for $i \simeq
90^o$).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          BENCHMARK MODEL
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\documentclass{article}
%% declare any packages used
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{color} 
\usepackage{pdfpages}
\usepackage{appendix}
\usepackage{subfigure}
\usepackage{amsmath}
\usepackage{amssymb}

\begin{document}

\title{Supplementary material for referee\\MNRAS: MN-15-0213-MJ}
\maketitle

\begin{figure}	%fullpage
\centering
\includegraphics[width=1.0\textwidth]{pops.png}
\caption{
}
\label{novalikes}
\end{figure}
